Let's talk about Thurstone \& Co.: An information-theoretical model for comparative judgments, and its statistical translation

### 2025-06-03 _tex/paper2.tex
Built site for gh-pages

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate, while \emph{estimates}
are the numerical values approximating the estimand
\citep{Neal_2020, Everitt_et_al_2010}.

@@

Following current CJ practices, researchers rely on estimates from the
BTL model--or its transformations--to approximate this estimand.
However, as discussed in Section~\ref{sec-theory-issues}, Thurstone's
Case V and the BTL model exhibit several statistical and measurement
limitations. These limitations hinder the model's ability to identify
various estimands relevant to CJ inquiries, including the one described
in the example.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different teaching methods influence students' writing
ability. Each structural component defines how observed or latent
variables affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. The model's hierarchical design allows
researchers to formulate and test hypotheses that account for both the
nested structure of stimuli and the uncertainties inherent in trait
estimation (see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for a discussion of these
considerations).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{\emph{Relevant variables} are
  those that satisfy the \emph{backdoor criterion} \citep[pp
  37]{Neal_2020}, that is, they belong to a \emph{sufficient adjustment
  set} \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} student-related
variables \(X_{I}\), such as teaching method, and students'
idiosyncratic errors \(e_{I}\) causally influence the latent variable
representing students' writing-quality trait \(T_{I}\). The error term
\(e_{I}\) captures variations in students' traits unexplained by
\(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows and
\(q_{I}\) independent columns (variables), and both \(e_{I}\) and
\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
  to footnote 2.} text-related variables \(X_{IA}\) (e.g., text length),
and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
written-quality trait \(T_{IA}\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) independent columns (variables), while \(e_{IA}\)
and \(T_{IA}\) are latent matrices with \(n_{I}\) rows and \(n_{A}\)
columns.

@@

Finally, we use \emph{Bayesian inference methods} to convert the
statistical model \ref{fig-cj16_stat} into a practical statistical tool
for analyzing paired comparison data. Bayesian inference offers three
main advantages in this context. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}. This
capability is essential for our implementation, as the model is indeed
overparameterized. Second, it incorporates prior information to
constrain parameter estimates within plausible bounds, thereby
mitigating estimation issues like non-convergence or improper solutions
that often affect frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. In our implementation, we
use mainly use priors to define the error distribution, setting the
scale of the latent variables \citep{Depaoli_2014}. Third, Bayesian
inference supports robust inferences from small samples, where the
asymptotic properties underlying frequentist methods are less reliable
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
feature is particularly relevant in CJ experiments, as researchers often
collect large volumes of paired comparisons but work with relatively
small samples of judges, stimuli, and individuals to test hypotheses.

@@

\subsection{Future research}\label{sec-discussion_RA}

Five current and new research avenues deserve particular attention due
to their direct impact on the reliability and validity of trait
estimates and the accuracy of statistical inferences. The following
sections outline these avenues and explain how our model provides a
rigorous research framework to investigate them.

@@

\subsubsection{The effects of non-dichotomous outcomes and
multidimensional traits on CJ experiments}\label{sec-discussion_RA5}

{About non-dichotomous outcomes and multidimensional traits}

\subsection{Challenges}\label{sec-discussion_challenges}

Despite its promise, this study presents several challenges that
researchers must navigate--challenges common to many quantitative
studies. They fall into two main categories. The first concerns the
foundational knowledge researchers need to apply the proposed framework
effectively. The second involves the increased attention researchers
must now devote to their specific research context.

@@

Researchers must develop foundational knowledge in two critical areas:
causal and Bayesian inference methods. One example that illustrates the
importance of mastering causal inference is our common assumption that
predictor variables are ``relevant'' to the research
context--interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this assumption entails or its implications for
model specification, estimation, and inference. As a result, to apply
the framework rigorously, researchers must engage deeply with these and
other complex ideas, such as how SCMs encode functional and
probabilistic information. To assist this effort, we have provided key
references throughout the study to guide motivated readers toward a
deeper understanding of these foundational concepts\footnote{refer to
  footnote 1, 2, and the detailed online document referenced in the
  Declarations section}.

Additionally, researchers must develop the expertise to translate
context-specific conceptual models into bespoke statistical models using
Bayesian inference methods. While Bayesian approaches introduce specific
technical demands--such as familiarity with probabilistic programming
languages (PPLs) and concepts like probability distributions and
convergence--we argue they offer a more accessible path for many applied
researchers. This is because Bayesian methods significantly reduce the
need for specialized knowledge in areas like optimization theory and
techniques, which are often necessary in frequentist approaches.

@@

When applying our framework to a specific CJ research context,
researchers face two additional challenges. First, they must verify that
the conceptual model provides a \emph{faithful} \footnote{Avoid
  confusing this term with the \emph{faithfulness assumption} described
  in the causal inference literature. For further details, see footnote
  1.} representation of the CJ system under study. Second, they must
evaluate whether the model's statistical translation can accurately
estimate the intended estimands (i.e., parameters) from empirical data.
Notably, addressing these challenges requires careful attention because
causal inference relies on expert knowledge and assumptions about the
variables' causal structure that are often untestable at the outset
\citep{Hernan_et_al_2025}. Consequently, the framework does not deliver
automatic answers; instead, it encourages researchers to formulate
precise questions and explicitly state their assumptions to build a
generalizable understanding of the phenomenon under study, in this case,
the CJ system \citep{Rohrer_et_al_2022, Deffner_et_al_2022}. This
theory-driven foundation then supports the empirical testing of causal
assumptions \citep{Deffner_et_al_2022}.

Thus, regarding the \emph{faithful} representation of the CJ system, we
advise researchers to treat our conceptual and statistical models as
starting points rather than universal solutions for all CJ designs or
datasets. This consideration means that researchers must face the
challenge of adapting these models to their specific research contexts,
paying close attention to causal assumptions, experimental design, and
model specification. This careful adaptation is essential because a
model's ability to produce accurate estimates and valid inferences
depends heavily on how well the data and inferential goals align with
its underlying assumptions \citep{Kohler_et_al_2019}. Nevertheless, as
\citet{Kohler_et_al_2019} points out, drawing conclusions about the real
world from observed data always requires assumptions, regardless of
whether the data are observational or experimental. In this context, our
approach provides a transparent framework for researchers to clearly
articulate these assumptions and design CJ experiments that support
reliable and valid trait measurement.

Finally, regarding the estimation capabilities of the statistical model,
researchers must tackle the challenge of performing identification
analysis. As outlined in Section~\ref{sec-theoretical},
\emph{identification analysis} determines whether a statistical model
can accurately compute a given estimand (e.g., a parameter) based solely
on its (causal) assumptions, independent of random variability
\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
necessary condition for consistency. \emph{Consistency} is the property
of an estimator (or model) whose estimates converge to the ``true''
value of an estimand as data size approaches infinity
\citep{Everitt_et_al_2010}. Logically, if researchers fail to identify a
model, it cannot achieve consistency--even with infinite, error-free
data. As a result, deriving meaningful insights from finite data becomes
impossible \citep{Schuessler_et_al_2023}.

Although performing identification analysis through formal derivations
may appear to be a natural next step, we argue that the complexity of
the CJ system renders such an approach impractical at the outset.
Instead, simulation-based methods, such as power analysis, provide a
more practical and flexible alternative, enabling researchers to examine
the consistency of estimates without relying on cumbersome mathematical
proofs. Nonetheless, the development of formal identification proofs
remains a significant goal for future CJ research.

>>>>

The \emph{structural approach} to causal inference provides a formal
framework for identifying causes and estimating their effects using
data. The approach uses structural causal models (SCMs) and directed
acyclic graphs (DAGs)
\citep{Pearl_2009, Pearl_et_al_2016, Gross_et_al_2018, Neal_2020} to
formally and graphically represent the assumed causal structure of a
system, such as the one found in CJ experiments. Essentially, SCMs and
DAGs function as \emph{conceptual models} on which identification
analysis rests. \emph{Identification analysis} determines whether an
estimator can accurately compute an estimand based solely on its
(causal) assumptions, regardless of random variability
\citep{Schuessler_et_al_2023}. Here, \emph{estimands} represent the
specific quantities researchers aim to determine (i.e., a parameter)
\citep{Everitt_et_al_2010}. \emph{Estimators} denote the methods or
functions that transform data into an estimate (e.g., a statistical
model), while \emph{estimates} are the numerical values approximating
the estimand \citep{Neal_2020, Everitt_et_al_2010}.

@@

Following standard CJ practices, researchers typically use estimates
from the BTL model--or its transformations--to approximate this
estimand. However, as discussed in Section~\ref{sec-theory-issues},
Thurstone's Case V and the BTL model exhibit several statistical and
measurement limitations. These limitations hinder the model's ability to
identify various estimands relevant to CJ inquiries, including the one
described in the example.

@@

Building on the principles of Structural Equation Modeling (SEM)
\citep{Hoyle_et_al_2023} and Item Response Theory (IRT)
\citep{Fox_2010, vanderLinden_et_al_2017_I}, the conceptual-population
model integrates two \emph{hierarchical structural components} to
examine how different \emph{relevant} \footnote{\emph{Relevant
  variables} are those that satisfy the \emph{backdoor criterion}
  \citep[pp 37]{Neal_2020}, that is, they belong to a \emph{sufficient
  adjustment set}
  \citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. A
  \emph{sufficient} set (potentially empty) blocks all non-causal paths
  between a predictor and an outcome without opening new ones
  \citep{Pearl_2009}. Refer also to footnote 1.} variables--whether
observed or latent--affect the primary latent variable of interest
\citep{Everitt_et_al_2010}. This hierarchical design enables researchers
to formulate and test hypotheses that account for both the nested
structure of stimuli and the uncertainties inherent in trait estimation
(see Section~\ref{sec-theory-issue1b} and
Section~\ref{sec-theory-issue2} for a discussion of these
considerations).

The top branch of DAG~\ref{fig-cj09_dag} illustrates the first
component, where \emph{relevant} \footnote{refer to footnote 2.}
student-related variables \(X_{I}\), such as teaching method, and
students' idiosyncratic errors \(e_{I}\) causally influence the latent
variable representing students' writing-quality trait \(T_{I}\). The
error term \(e_{I}\) captures variations in students' traits unexplained
by \(X_{I}\). Here, \(X_{I}\) is an observed matrix with \(n_{I}\) rows
and \(q_{I}\) independent columns (variables), and both \(e_{I}\) and
\(T_{I}\) are latent vectors of length \(n_{I}\). Additionally, this
branch shows how \(T_{I}\), along with \emph{relevant} \footnote{refer
  to footnote 2.} text-related variables \(X_{IA}\) (e.g., text length),
and texts' idiosyncratic errors \(e_{IA}\) causally influence the texts'
written-quality trait \(T_{IA}\), the first primary latent variable of
interest. The error term \(e_{IA}\) captures variations in the texts'
traits that remain unexplained by \(T_{I}\) or \(X_{IA}\). Here,
\(X_{IA}\) is an observed matrix with dimensions \(n_{I} \cdot n_{A}\)
rows and \(q_{IA}\) independent columns (variables), while \(e_{IA}\)
and \(T_{IA}\) are latent matrices with \(n_{I}\) rows and \(n_{A}\)
columns.

@@

Finally, we use \emph{Bayesian inference methods} to convert the
statistical model \ref{fig-cj16_stat} into a practical statistical tool
for analyzing paired comparison data. Bayesian inference offers three
main advantages in this context. First, it handles complex and
overparameterized models, where the number of parameters exceeds the
number of observations \citep{Baker_1998, Kim_et_al_1999}. This feature
is essential for our implementation, as the proposed model is indeed
overparameterized. Second, it incorporates prior information to
constrain parameter estimates within plausible bounds, thereby
mitigating estimation issues like non-convergence or improper solutions
that often affect frequentist methods
\citep{Martin_et_al_1975, Seaman_et_al_2011}. We primarily use this
capability to define the error distribution and set the scale of latent
variables \citep{Depaoli_2014}. Third, Bayesian inference supports
robust inferences from small samples, where the asymptotic properties
underlying frequentist methods are less reliable
\citep{Baldwin_et_al_2013, Lambert_et_al_2006, Depaoli_2014}. This
feature is particularly relevant in CJ experiments, as researchers often
collect large volumes of paired comparisons but work with relatively
small samples of judges, stimuli, and individuals to test hypotheses.

@@

\subsection{Further research}\label{sec-discussion_RA}

Four current and new research avenues deserve particular attention due
to their direct impact on the reliability and validity of trait
estimates and the accuracy of statistical inferences. The following
sections outline these avenues and explain how our model provides a
rigorous research framework to investigate them.

@@

\subsection{Limitations and challenges}\label{sec-discussion_challenges}

Drawing conclusions about the real world from observed data always
requires assumptions, regardless of whether the data are observational
or experimental \citep{Kohler_et_al_2019, Deffner_et_al_2022}. This
study is no exception to this principle. By adopting causal inference as
the theoretical backbone of our proposed framework, we necessarily rely
on expert knowledge and assumptions about the variables' causal
structure that are often untestable at the outset
\citep{Hernan_et_al_2025}. As such, our framework does not yield
automatic answers. Instead, it encourages researchers to formulate
precise questions and to explicitly articulate their assumptions to
develop a generalizable understanding of the CJ system
\citep{Rohrer_et_al_2022, Deffner_et_al_2022, Sterner_et_al_2024}. This
clarity is essential because a model's ability to produce accurate
estimates and valid inferences depends heavily on how well the data and
inferential goals align with its underlying assumptions
\citep{Kohler_et_al_2019}--a property we cannot yet guarantee for our
models. However, this theory-driven approach offers a key benefit: its
lays the foundation for empirically testing the proposed causal
assumptions \citep{Deffner_et_al_2022}.

In light of these limitation, this study presents several challenges
that researchers must navigate--challenges common to many quantitative
studies. These challenges fall into two main categories. The first
concerns the foundational knowledge researchers need to apply our
proposed framework effectively. The second involves the increased
attention researchers must now devote to their specific research
context.

@@

In terms of foundational knowledge, researchers must face two core
challenges to use our framework effectively. First, they need a solid
understanding of causal inference principles. Second, they must learn to
translate the functional and probabilistic content of conceptual models
into bespoke statistical models. One example that illustrates the
importance of a sound knowledge of causal inference is the recurrent
assumption that predictor variables are ``relevant'' to the research
context---interpreting this relevance as their inclusion in a
\emph{sufficient adjustment set}
\citep{Pearl_2009, Pearl_et_al_2016, Morgan_et_al_2014}. However, we do
not fully explore what this assumption entails or its implications for
model specification, estimation, and inference. As a result, researchers
must engage deeply with these and other complex ideas, such as how SCMs
encode functional and probabilistic information. To assist this effort,
we include key references throughout the study to guide motivated
readers toward deeper understanding\footnote{refer to footnote 1, 2, and
  the detailed online document referenced in the Declarations section}.

Moreover, developing the skills to translate functional and
probabilistic information into bespoke statistical models has its
challenges. Although we argue that Bayesian inference methods offer a
more accessible path for many applied researchers to develop these
skills--since they reduce the need for specialized knowledge in areas
like optimization theory, which frequentist methods often require--these
methods still demand familiarity with technical concepts, such as
probabilistic programming languages (PPLs), probability distributions,
and convergence, that researchers must acquire. Nevertheless, to support
researchers in building these skills, we include links to model code and
alternative specifications in the \textbf{Declarations} section of this
document.

@@

Regarding attention to the research context, researchers face two
additional challenges. First, they must verify that the conceptual model
provides a \emph{faithful} \footnote{Avoid confusing this term with the
  \emph{faithfulness assumption} described in the causal inference
  literature. For further details, see footnote 1.} representation of
the CJ system under study. Second, they must assess whether the
statistical translation of the model can accurately estimate the
intended estimands (i.e., parameters) from empirical data. To ensure a
faithful representation of the CJ system, we encourage researchers to
treat our conceptual and statistical models as starting points rather
than universal solutions for all CJ designs or datasets. As such,
researchers must tackle the challenge of adapting these models to their
specific research contexts, paying close attention to causal
assumptions, experimental design, and model specification. Nonetheless,
our approach offers a transparent framework that enables researchers to
explicitly articulate any new assumptions and design CJ experiments that
support reliable and valid trait measurement.

Conversely, to evaluate the estimation capabilities of the statistical
model, researchers must tackle the challenge of performing
identification analysis. As outlined in Section~\ref{sec-theoretical},
\emph{identification analysis} determines whether a statistical model
can accurately compute a given estimand (e.g., a parameter) based solely
on its (causal) assumptions, independent of random variability
\citep{Schuessler_et_al_2023}. Identification is crucial because it is a
necessary condition for consistency. \emph{Consistency} is the property
of an estimator (e.g., a statistical model) whose estimates converge to
the ``true'' value of an estimand as data size approaches infinity
\citep{Everitt_et_al_2010}. Without identification, consistency is
impossible--even with infinite, error-free data--and thus, meaningful
inference from finite samples cannot be achieved
\citep{Schuessler_et_al_2023}.

Although performing identification analysis through formal derivations
may appear to be a natural next step, we argue that the complexity of
the CJ system renders such an approach impractical at the outset.
Instead, simulation-based methods, such as power analysis, provide a
more practical and flexible alternative, enabling researchers to examine
the consistency of estimates without relying on cumbersome mathematical
proofs. Nonetheless, the development of formal identification proofs
remains a significant goal for future CJ research. Notably, our
framework supports both approaches, providing a foundation for formal
derivations as well as simulation-based methods.
